# Week 2

*This week's exercises were about data wrangling and performing* 
*and interpreting a simple linear regression*


```{r}
date()
```

Read in the students2014 data

```{r}
learning14 <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/learning2014.txt", header = TRUE, sep = ",")
```

```{r}
str(learning14)
```
The data set consists of 7 variables and 166 observations. Gender is a character variable,
and the others are integers. -> changing gender to factor variable

```{r}
learning14$gender <- as.factor(learning14$gender)
summary(learning14)
```
Information on the variables is available from https://www.mv.helsinki.fi/home/kvehkala/JYTmooc/JYTOPKYS2-meta.txt. 

Short description

Age: Age (in years) derived from the date of birth

Gender: Male = 1  Female = 2

Attitude: Global attitude toward statistic

Deep:  Deep approach

Surf: Surface approach

Stra: Strategic approach   

Points: Exam points

It seems that majority of students are female (wondering whether non-binary was an option...), 
they are aged 17-55, 
and they have received exam points between 7 and 33 (mean 22.7) from 
Introduction to Social Statistics course in fall 2014. 


```{r}
library(reshape2)
#Convert wide to long
l14.melt <- melt(learning14)  # gender as id
head(l14.melt)
```
# Graphical overview

```{r}
library(ggplot2)
ggplot(data = l14.melt, aes(x = value, fill = gender)) + 
  stat_density(alpha = 0.5, position = "identity") + facet_wrap(~variable, scales = "free")
```

Plot count histograms as comparison (to see if we have enough data for smoothing distribution plots as done above).
(Non-stacking and alpha to see through.)

```{r}
ggplot(data = l14.melt, aes(x = value, fill = gender)) +
        geom_histogram(aes(y = ..count..), position = "identity", alpha = 0.5) + facet_wrap(~variable, scales = "free")
```

Some findings based on above graphs:

- the age distribution is skewed towards younger, most students are in their early 20s
- female students have lower attitude points compared to male students
- female students have somewhat higher points in strategic approach (stra)


### Correlations

```{r}
#install.packages('corrplot')
library(corrplot)
library(RColorBrewer)
newdata <- learning14
newdata$gender <- as.numeric(newdata$gender)
#M <-cor(newdata)
#corrplot(M, type="upper", order="hclust",
#         col=brewer.pal(n=8, name="RdYlBu"))

source("http://www.sthda.com/upload/rquery_cormat.r")
rquery.cormat(newdata)
```
The strongest positive correlation is between exam points and attitude. 
Gender also seems to be correlated with attitude.
The strongest negative correlation is between deep and surface approaches.

# Linear regression

Let's fit a linear model with age, gender and attitude es explanatory variables and 
exam points as an outcome


```{r}
model1 <- lm(points~ age + gender + attitude, data = learning14)
summary(model1)
```
## Interpretation

Attitude is positively associated with exam points the estimate being 3.6
(when attitude increases by one, points increase by 3.6) 
The association is statistically significant (p-value < 0.001). 
Age and gender have negative, non-significant association. -> Let's remove them from the model.
The estimate of attitude decreases a bit (3.5) and remains strongly significant. 

Multiple R-squared of the model with all three explanatory variables is 0.2018 
which means that about 20% of the variation in exam points can be explained by the three variables.

## Model diagnostics

Testing assumptions of linear model.

### Residuals vs fitted values

```{r}
plot(model1, 1)
```

There seems not to be a pattern in the red line, suggesting it is safe to assume 
linear relationship between the explanatory variables and exam points outcome.
 
### Normality of residuals

```{r}
plot(model1, 2)
```

Based on QQ-plot there seems not be any strong violation to the normality assumption as there is no clear pattern. 

### Residuals vs Leverage

```{r}
plot(model1, 5)
```

Residuals vs leverage plot shows three outliers, these could be explored further. 
